---
title: "Whole brain clustering"
output: 
  html_notebook: 
    toc: true
    theme: cerulean
    number_sections: true
---

# . Environment
## Load libraries
```{r}
#convert from .h5ad
library(zellkonverter)
library(SingleCellExperiment)
library(Matrix)

# Seurat
library(Seurat)
library(clustree)

# General
library(dplyr)
library(ggplot2)
```

## define functions
### Run binomial glm
```{r}
run_binomial_glm <- function(seurat_obj, group_var) {
  library(dplyr)
  library(tidyr)
  library(purrr)
  library(broom)
  
  # Extract metadata
  df <- seurat_obj@meta.data
  
  # Ensure required columns exist
  if (!all(c("population", "sample_id", group_var) %in% colnames(df))) {
    stop("Metadata must contain: population, sample_id, and the specified group_var.")
  }
  
  # Relevel population and factorize inputs
  df <- df %>%
    mutate(
      population = relevel(factor(population), ref = "surface"),
      sample_id  = factor(sample_id),
      !!group_var := factor(.data[[group_var]])
    )
  
  # Summarize counts per (sample × group_var)
  sample_sums <- df %>%
    group_by(sample_id, population, !!sym(group_var)) %>%  # sym() avoids tidyselect
    summarise(count = n(), .groups = "drop")
  
  # Total per sample
  total_by_sample <- sample_sums %>%
    group_by(sample_id) %>%
    summarise(total = sum(count), .groups = "drop")
  
  # Add "non" counts
  sample_sums <- sample_sums %>%
    left_join(total_by_sample, by = "sample_id") %>%
    mutate(non = total - count)
  
  # GLM function
  fit_binom <- function(df_sub) {
    glm(cbind(count, non) ~ population,
        family = binomial,
        data = df_sub)
  }
  
  # Fit per group_var level
  glm_results <- sample_sums %>%
    rename(group = !!sym(group_var)) %>%    # safe non-tidyselect
    group_by(group) %>%
    nest() %>%
    mutate(
      model = map(data, fit_binom),
      tidy  = map(model, broom::tidy)
    ) %>%
    select(group, tidy) %>%
    unnest(tidy)
  
  # Final cleaned results
   glm_effects <- glm_results %>%
    filter(term != "(Intercept)") %>%
    mutate(
      odds_ratio = exp(estimate),
      p_adj_num  = p.adjust(p.value, method = "fdr"),
      p_adj      = format.pval(p_adj_num, digits = 3, eps = 1e-300)
    ) %>%
    select(-p_adj_num) %>%
    arrange(p.value)
  
  return(glm_effects)
}
```
### Annotation - DEGs across cell types
```{r}
deseq2_pb_annotation <- function(
  cts,
  coldata,
  target_celltype,
  ref_celltypes = NULL,
  fdr_explore = 0.20,
  min_total_count = 10,
  min_samples = 2
) {

  stopifnot(is.matrix(cts) || inherits(cts, "dgCMatrix"))
  stopifnot(all(c("col_id","sample","Population","CellType") %in% colnames(coldata)))

  # clean coldata
  coldata <- as.data.frame(coldata, stringsAsFactors = FALSE)
  coldata$col_id     <- as.character(coldata$col_id)
  coldata$sample     <- as.character(coldata$sample)
  coldata$Population <- as.character(coldata$Population)
  coldata$CellType   <- as.character(coldata$CellType)

  # USE BOTH POPULATIONS for annotation (no subsetting)
  cd <- coldata
  if (nrow(cd) == 0) stop("coldata is empty after cleaning.")

  # optionally restrict which “other” celltypes are allowed
  if (!is.null(ref_celltypes)) {
    keep_cts <- unique(c(ref_celltypes, target_celltype))
    cd <- cd[cd$CellType %in% keep_cts, , drop = FALSE]
    if (nrow(cd) == 0) stop("After ref_celltypes filtering, nothing left.")
  }

  # need at least some target samples
  if (!any(cd$CellType == target_celltype)) {
    stop("No pseudobulk columns for target celltype: ", target_celltype)
  }

  # align counts to cd
  stopifnot(all(cd$col_id %in% colnames(cts)))
  cts2 <- cts[, cd$col_id, drop = FALSE]

  # define target vs other
  cd$is_target <- ifelse(cd$CellType == target_celltype, "target", "other")
  cd$is_target <- factor(cd$is_target, levels = c("other","target"))

  # Prefilter genes: expressed in at least min_samples pseudobulk columns at >= min_total_count
  keep <- Matrix::rowSums(cts2 >= min_total_count) >= min_samples
  cts2 <- cts2[keep, , drop = FALSE]
  if (nrow(cts2) == 0) stop("All genes filtered out; relax thresholds.")

  # DESeq2 colData must match column order
  cd2 <- cd[match(colnames(cts2), cd$col_id), , drop = FALSE]
  stopifnot(identical(colnames(cts2), cd2$col_id))
  rownames(cd2) <- cd2$col_id

  # IMPORTANT: sample as a blocking factor (paired within sample)
  cd2$sample    <- factor(cd2$sample)
  cd2$is_target <- factor(cd2$is_target, levels = c("other","target"))

  dds <- DESeq2::DESeqDataSetFromMatrix(
    countData = round(cts2),
    colData   = cd2,
    design    = ~ sample + is_target
  )

  # sanity: need both target and other
  if (length(unique(cd2$is_target)) < 2) {
    stop("Need both target and other columns after filtering.")
  }

  dds <- DESeq2::DESeq(dds)

  # results for target vs other
  res <- DESeq2::results(
    dds,
    contrast = c("is_target","target","other"),
    alpha = fdr_explore
  ) |>
    as.data.frame() |>
    tibble::rownames_to_column("gene")

  res_all <- res |>
    dplyr::arrange(padj, dplyr::desc(baseMean))

  res_sig <- res_all |>
    dplyr::filter(!is.na(padj), padj < fdr_explore)

  list(
    dds             = dds,
    res_sig         = res_sig,
    res_all         = res_all,
    target_celltype = target_celltype,
    fdr_explore     = fdr_explore
  )
}
```
##### Export single cell type significant DEGs to csv
```{r}
format_and_write_annotation_csv <- function(
  res_df,
  neuro_cluster_label,          # e.g. "e2_tectum"
  padj_cut = 0.05,
  lfc_abs_min = 0,
  top_n = 100,
  drop_LOC = TRUE,
  loc_regex = "^LOC",
  extra_drop_regex = NULL
) {

  stopifnot(all(c("gene","padj","baseMean") %in% colnames(res_df)))

  df <- res_df

  # Optional filtering of LOC / unwanted symbols
  if (drop_LOC) {
    df <- df %>% filter(!grepl(loc_regex, gene))
  }
  if (!is.null(extra_drop_regex)) {
    df <- df %>% filter(!grepl(extra_drop_regex, gene))
  }

  # Keep significant genes
  df <- df %>% filter(!is.na(padj), padj <= padj_cut)

  # Determine which LFC column to use
  lfc_col <- if ("log2FoldChange_shrunk" %in% colnames(df)) {
    "log2FoldChange_shrunk"
  } else if ("log2FoldChange" %in% colnames(df)) {
    "log2FoldChange"
  } else {
    stop("res_df must contain log2FoldChange or log2FoldChange_shrunk")
  }

  # Keep only TARGET-ENRICHED genes
  df <- df %>% filter(.data[[lfc_col]] > 0)

  # Optional LFC threshold
  if (lfc_abs_min > 0) {
    df <- df %>% filter(abs(.data[[lfc_col]]) >= lfc_abs_min)
  }

  # Add direction column (useful for sanity checks)
  df <- df %>%
    mutate(direction = "Up in target")

  # Rank by: significance → specificity → expression
  df_out <- df %>%
    arrange(padj, desc(abs(.data[[lfc_col]])), desc(baseMean)) %>%
    slice_head(n = top_n)

  # Column order (clean CSV)
  preferred <- c("gene","baseMean",lfc_col,"padj","pvalue","stat","lfcSE","direction")
  keep_cols <- c(preferred, setdiff(colnames(df_out), preferred))
  df_out <- df_out %>% select(any_of(keep_cols))

  # Write directly to WORKING DIRECTORY
  fname <- paste0(neuro_cluster_label, ".csv")
  readr::write_csv(df_out, fname)

  message("Wrote annotation DEG CSV to: ", normalizePath(fname))

  return(df_out)
}
```

##### Export all cell type markers with log2fold change cutoff
```{r}
run_all_celltype_deg_exports <- function(
  cts,
  coldata,
  padj_cut = 0.05,
  lfc_min = 1,
  positive_only = TRUE,
  out_csv = "AllCellTypes_DEGs_filtered.csv",
  verbose = TRUE
) {
  suppressPackageStartupMessages({
    library(dplyr)
    library(readr)
  })

  stopifnot(is.matrix(cts) || inherits(cts, "dgCMatrix"))
  stopifnot(all(c("col_id","sample","Population","CellType") %in% colnames(coldata)))

  cd <- as.data.frame(coldata, stringsAsFactors = FALSE)
  celltypes <- sort(unique(cd$CellType))
  celltypes <- celltypes[!is.na(celltypes) & celltypes != ""]

  if (length(celltypes) == 0) stop("No celltypes found in coldata.")

  results_list <- vector("list", length(celltypes))
  names(results_list) <- celltypes

  failures <- list()

  for (ct in celltypes) {
    if (isTRUE(verbose)) message("Running DESeq2 for: ", ct)

    out <- try(
      deseq2_pb_annotation(
        cts = cts,
        coldata = cd,
        target_celltype = ct
      ),
      silent = TRUE
    )

    if (inherits(out, "try-error")) {
      failures[[ct]] <- as.character(out)
      next
    }

    df <- out$res_all %>%
      mutate(celltype = ct) %>%
      select(celltype, gene, padj, log2FoldChange, lfcSE, pvalue, baseMean) %>%
      filter(!is.na(padj), padj <= padj_cut) %>%
      filter(!is.na(log2FoldChange))

    if (positive_only) {
      df <- df %>%
        filter(log2FoldChange >= lfc_min)
    } else {
      df <- df %>%
        filter(abs(log2FoldChange) >= lfc_min)
    }

    df <- df %>%
      arrange(celltype, desc(log2FoldChange), padj, desc(baseMean))

    results_list[[ct]] <- df
  }

  combined_results <- bind_rows(results_list)

  readr::write_csv(combined_results, out_csv)

  if (isTRUE(verbose)) {
    message("Wrote filtered DEG file: ", normalizePath(out_csv))
    if (length(failures) > 0) {
      message("Some cell types failed. Check returned $failures.")
    }
  }

  return(list(
    degs = combined_results,
    failures = failures
  ))
}
```

### Plotting
#### UMAP arrows for coord axis
```{r}
mini_axis_umap <- function(
  seurat_obj,
  reduction = "umap",
  group.by = NULL,
  split.by = NULL,
  label = TRUE,
  shuffle = TRUE,
  alpha = 0.8,
  pt.size = 0.5,
  axis_frac = 0.08,
  pad_frac  = 0.03,
  arrow_mm  = 2,
  linewidth = 0.5,
  add_labels = TRUE,
  label_x = "UMAP 1",
  label_y = "UMAP 2",
  label_size = 3,
  margin_mm = 5.5,
  seed = NULL
) {
  stopifnot(inherits(seurat_obj, "Seurat"))

  if (!is.null(seed)) set.seed(seed)

  # 1) Base DimPlot
  p <- DimPlot(
    seurat_obj,
    reduction = reduction,
    group.by = group.by,
    split.by = split.by,
    label = label,
    shuffle = shuffle,
    alpha = alpha,
    pt.size = pt.size
  ) + NoLegend()

  # 2) Remove axes completely
  p_clean <- p +
    theme_void() +
    theme(plot.margin = margin(margin_mm, margin_mm, margin_mm, margin_mm, unit = "mm")) +
    NoLegend()

  # 3) Compute panel ranges
  gb <- ggplot_build(p_clean)
  pp <- gb$layout$panel_params[[1]]
  x_rng <- pp$x.range
  y_rng <- pp$y.range

  x_span <- diff(x_rng)
  y_span <- diff(y_rng)

  # Bottom-left corner position (with padding)
  x0 <- x_rng[1] + pad_frac * x_span
  y0 <- y_rng[1] + pad_frac * y_span

  # Mini-axis lengths
  dx <- axis_frac * x_span
  dy <- axis_frac * y_span

  # 4) Add mini-axes + optional labels
  p_final <- p_clean +
    annotate(
      "segment",
      x = x0, xend = x0 + dx,
      y = y0, yend = y0,
      linewidth = linewidth,
      arrow = arrow(type = "closed", length = unit(arrow_mm, "mm"))
    ) +
    annotate(
      "segment",
      x = x0, xend = x0,
      y = y0, yend = y0 + dy,
      linewidth = linewidth,
      arrow = arrow(type = "closed", length = unit(arrow_mm, "mm"))
    )

  if (isTRUE(add_labels)) {
    p_final <- p_final +
      annotate(
        "text",
        x = x0 + dx, y = y0 - 0.01 * y_span,
        label = label_x, hjust = 1, vjust = 1, size = label_size
      ) +
      annotate(
        "text",
        x = x0 - 0.01 * x_span, y = y0 + dy,
        label = label_y, hjust = 1, vjust = 1, angle = 90, size = label_size
      )
  }

  return(p_final)
}
```

# . Data
skip to load data - conversion to lSeq from python output included for transparency
## read in python files 
```{r}
# sce_counts <- readH5AD("raw_counts.h5ad")
# sce_lognorm <- readH5AD("initial_clusters_all_resolutions_RAWasX.h5ad")
# sce_scaled  <- readH5AD("initial_clusters_all_resolutions.h5ad")
```
### change dim names in scaled data and extract highest resolution leiden clustering
```{r}
# if (!is.null(sce_scaled@int_colData$reducedDims)) {
#   if (!is.null(sce_scaled@int_colData$reducedDims$X_pca)) {
#     reducedDim(sce_scaled, "X_pca") <- sce_scaled@int_colData$reducedDims$X_pca
#   }
#   if (!is.null(sce_scaled@int_colData$reducedDims$X_pca_harmony)) {
#     reducedDim(sce_scaled, "X_pca_harmony") <- sce_scaled@int_colData$reducedDims$X_pca_harmony
#   }
#   if (!is.null(sce_scaled@int_colData$reducedDims$X_umap)) {
#     reducedDim(sce_scaled, "X_umap") <- sce_scaled@int_colData$reducedDims$X_umap
#   }
# }
# 
# #Fix column names for each embedding matrix
# if ("X_pca" %in% reducedDimNames(sce_scaled)) {
#   colnames(reducedDim(sce_scaled, "X_pca")) <- paste0("PC_", seq_len(ncol(reducedDim(sce_scaled, "X_pca"))))
# }
# if ("X_pca_harmony" %in% reducedDimNames(sce_scaled)) {
#   colnames(reducedDim(sce_scaled, "X_pca_harmony")) <- paste0("Harmony_", seq_len(ncol(reducedDim(sce_scaled, "X_pca_harmony"))))
# }
# if ("X_umap" %in% reducedDimNames(sce_scaled)) {
#   colnames(reducedDim(sce_scaled, "X_umap")) <- paste0("UMAP_", seq_len(ncol(reducedDim(sce_scaled, "X_umap"))))
# }
# 
# #Rename them to match Seurat-friendly names
# if ("X_pca" %in% reducedDimNames(sce_scaled)) {
#   reducedDim(sce_scaled, "pca") <- reducedDim(sce_scaled, "X_pca")
# }
# if ("X_pca_harmony" %in% reducedDimNames(sce_scaled)) {
#   reducedDim(sce_scaled, "harmony") <- reducedDim(sce_scaled, "X_pca_harmony")
# }
# if ("X_umap" %in% reducedDimNames(sce_scaled)) {
#   reducedDim(sce_scaled, "umap") <- reducedDim(sce_scaled, "X_umap")
# }
# 
# # remove old ones to avoid confusion
# keep_rd <- intersect(c("pca", "harmony", "umap"), reducedDimNames(sce_scaled))
# if (length(keep_rd) > 0) {
#   reducedDims(sce_scaled) <- reducedDims(sce_scaled)[keep_rd]
# }
# 
# # -----------------------------
# # Subset to leiden_1.20 cells
# # -----------------------------
# if (!"leiden_1.20" %in% colnames(colData(sce_scaled))) stop("sce_scaled is missing 'leiden_1.20' in colData.")
# 
# cells_120 <- colnames(sce_scaled)  # all cells in this object already have leiden labels
# sce_counts  <- sce_counts[,  cells_120]
# sce_lognorm <- sce_lognorm[, cells_120]
# sce_scaled  <- sce_scaled[,  cells_120]
# 
# stopifnot(
#   identical(colnames(sce_counts),  colnames(sce_scaled)),
#   identical(colnames(sce_lognorm), colnames(sce_scaled))
# )

```

### Convert to seurat object
#### Build Seurat object from raw counts matrix 
```{r}
# counts_mat <- assay(sce_counts, "X")
# if (!inherits(counts_mat, "dgCMatrix")) counts_mat <- as(counts_mat, "dgCMatrix")
# 
# meta <- as.data.frame(colData(sce_scaled)[, c("sample_id", "population", "leiden_1.20"), drop = FALSE])
# 
# seu <- CreateSeuratObject(
#   counts = counts_mat,
#   assay  = "RNA",
#   meta.data = meta
# )
# 
# DefaultAssay(seu) <- "RNA"
```

#### Add normalization to Data
```{r}
# lognorm_mat <- assay(sce_lognorm, "X")
# lognorm_mat <- lognorm_mat[, colnames(seu), drop = FALSE]
# 
# seu <- SetAssayData(
#   object  = seu,
#   assay   = "RNA",
#   layer   = "data",
#   new.data = lognorm_mat
# )
```

#### Add scaled highly variable features
```{r}
# scaled_mat <- assay(sce_scaled, "X")
# scaled_mat <- scaled_mat[, colnames(seu), drop = FALSE]
# 
# hvg <- intersect(rownames(scaled_mat), rownames(seu))
# if (length(hvg) == 0) stop("No overlapping HVG features between scaled and Seurat features.")
# 
# scaled_mat <- scaled_mat[hvg, , drop = FALSE]
# VariableFeatures(seu) <- hvg
# 
# seu <- SetAssayData(
#   object  = seu,
#   assay   = "RNA",
#   layer   = "scale.data",
#   new.data = scaled_mat
# )
```

#### Add reductions from SCE (pca/harmony/umap)
```{r}
# add_rd <- function(seu, sce, rd, key) {
#   if (!rd %in% reducedDimNames(sce)) return(seu)
#   emb <- reducedDim(sce, rd)
#   emb <- emb[colnames(seu), , drop = FALSE]
#   # Keep existing colnames (we already set them), but ensure they exist
#   if (is.null(colnames(emb))) colnames(emb) <- paste0(key, seq_len(ncol(emb)))
#   seu[[rd]] <- CreateDimReducObject(embeddings = emb, key = key, assay = "RNA")
#   seu
# }
# 
# seu <- add_rd(seu, sce_scaled, "pca",     "PC_")
# seu <- add_rd(seu, sce_scaled, "harmony", "Harmony_")
# seu <- add_rd(seu, sce_scaled, "umap",    "UMAP_")
# 
# #  Leiden -> seurat_clusters 
# seu$seurat_clusters <- factor(seu$leiden_1.20)
# Idents(seu) <- "seurat_clusters"
```

## save
```{r}
# lSeq <- seu
 SaveSeuratRds(lSeq, "lSeq.rds")
```

## load
```{r}
lSeq <-readRDS("lSeq.rds")
```


## QC plots
### UMAPs for integration
POPULATION
```{r}
lSeq$population <- factor(lSeq$population, levels = c("surface", "cave"))

DimPlot(lSeq, reduction = "umap", group.by = "population", label = FALSE,pt.size = 0.25,
   cols = c("darkseagreen","palevioletred"))

#ggsave("lSeqPopulationUMAP.pdf",  width = 10, height = 8, dpi = 300)

```
SAMPLE
```{r}

DimPlot(lSeq, reduction = "umap", group.by = "sample_id", label = FALSE,pt.size = 0.25)

#ggsave("lSeqSampleUMAP.pdf",  width = 10, height = 8, dpi = 300)

```
### QC cutoffs plots
```{r}
Idents(lSeq) <- lSeq$sample_id
```

#### ribos
```{r}
#create column of ribosomal gene expression percentage
lSeq[["percent.ribo"]] <- PercentageFeatureSet(
  lSeq, 
  pattern = "^[Rr][Pp][LlSs]"
)
```
```{r}

VlnPlot(
  lSeq,
  features = "percent.ribo",
  pt.size = 0,
  adjust = 5
)  + NoLegend()


ggsave("lSeqRiboViolin.pdf",  width = 10, height = 8, dpi = 300)

```

```{r}
VlnPlot(
  lSeq,
  features = "percent.ribo",
  group.by = "population",
  cols = c("darkseagreen","palevioletred"),
  pt.size = 0,
   adjust = 5
) + NoLegend()

#ggsave("Ribo_by_population_violin.pdf", width = 6, height = 4, dpi = 300)
```

#### genes
```{r}
VlnPlot(lSeq, features = c("nFeature_RNA"), pt.size = 0,  adjust = 5)  + NoLegend()
#ggsave("lSeqGenesViolin.pdf",  width = 10, height = 8, dpi = 300)
```
```{r}
VlnPlot(
  lSeq,
  features = "nFeature_RNA",
  group.by = "population",
  cols = c("darkseagreen","palevioletred"),
  pt.size = 0
) + NoLegend()

#ggsave("Genes_by_population_violin.pdf", width = 6, height = 4, dpi = 300)
```

#### UMIs
```{r}
VlnPlot(lSeq, features = c("nCount_RNA"), pt.size = 0,  adjust = 5)  + NoLegend()
#ggsave("lSeqUMIViolin.pdf",  width = 10, height = 8, dpi = 300)
```

```{r}
VlnPlot(
  lSeq,
  features = "nCount_RNA",
  group.by = "population",
  cols = c("darkseagreen","palevioletred"),
  pt.size = 0
) + NoLegend()

#ggsave("UMI_by_population_violin.pdf", width = 6, height = 4, dpi = 300)
```

## properties
### cell count totals
```{r}
# total cells
n_total_cells <- ncol(lSeq)

# cells per population
cells_by_pop <- table(lSeq$population)

n_total_cells
cells_by_pop
```
### UMIs
#### total
```{r}
mean_umis <- mean(lSeq$nCount_RNA)
median_umis <- median(lSeq$nCount_RNA)

mean_umis
median_umis
```

#### by pop
```{r}
umis_by_pop <- lSeq@meta.data %>%
  group_by(population) %>%
  summarize(
    mean_umis   = mean(nCount_RNA),
    median_umis = median(nCount_RNA),
    .groups = "drop"
  )

umis_by_pop
```
#### stats
sample level - surface and cave populations counts are not significantly different across samples
```{r}
sample_qc <- lSeq@meta.data %>%
  group_by(sample_id, population) %>%
  summarize(
    median_umis = median(nCount_RNA),
    mean_umis   = mean(nCount_RNA),
    .groups = "drop"
  )

sample_qc

wilcox.test(median_umis ~ population, data = sample_qc)
```

```{r}
p_val <- wilcox.test(median_umis ~ population, data = sample_qc)$p.value

ggplot(sample_qc, aes(x = population, y = median_umis)) +
  geom_violin(trim = FALSE) +
  geom_jitter(width = 0.1, size = 2) +
  theme_classic() +
  annotate(
    "text",
    x = 1.5,
    y = max(sample_qc$median_umis) * 1.05,
    label = paste0("Wilcoxon p = ", signif(p_val, 2))
  ) +
  labs(
    x = NULL,
    y = "Median UMIs per cell (per sample)"
  )
```

### genes
```{r}
median(lSeq$nFeature_RNA)
```

# . Annotate by major cell type group
Clusters initially annotated as neurons, glia, or other
## Qick sort by module score
### Gene list
LOC125803969 = mrc1b
```{r}
neuronal_markers <- c(
  "elavl2","elavl3", "elavl4", "rbfox3a", "snap25a",
  "slc17a6a", "slc17a6b", "slc17a7a", "slc17a7b", "slc17a8",
  "gad1a", "gad1b", "gad2",
  "slc6a5", "slc32a1", "sv2a", "vamp2",
  "mao", "ddc", "tph1a", "tph1b", "tph2",
  "th", "slc6a3", "th2", "slc18a2", "hrh3",
  "chata", "slc18a3a", "slc18a3b", "oxt", "hcrt"
)

glial_markers <- c(
  # Astrocytes & radial glia
  "fabp7a", "LOC103040661", "her9", "slc1a2b", "s100b",
  "sox2", "LOC103035120", "slc1a3b", "glula",
  "kcnj10a", "slc6a11b", "fgfr3", "fgfr4", "ptn",
  "aqp4", "slc1a2b", "mlc1", "ptprz1a", 
  "dla", "ccnd1", "mki67", "her2", "uhrf1",
  "pcna", "stil", "cdca3", "cdca7a", "s100b",
  
  # Oligodendrocyte lineage
  "olig2", "sox10", "plp1b", "pdgfra", "col9a3",
  "matn4", "mbpa", "bcan", "cntn1b",
  "olig1", "myt1b", "cspg4", "ptprfa",
  "lingo1a", "lingo1b", "plppr2b", "mbpb", "plp1a",
  "mpz", "mag", "cldnk", "cldn19", "gldn", "cd9b",
  "myrf", "qki2", "bcas1", "rundc3aa",
  "sall1b", "enpp6", "ugt8", "gal3st1a",
  
  # Microglia/macrophages
  "irf1a", "stat4", "parp9", "socs3b", "il12rb2",
  "mavs", "ptprc", "ackr4b", "coro1a", "fadd",
  "ccl44", "ccl25a", "csf1rb", "havcr2",
  "il1rl1", "tlr3", "irf1a", "stat4", "parp9",
  "socs3b", "il12rb2", "mavs", "ptprc", "ackr4b",
  "coro1a", "fadd", "ccl44", 
  "tmem119a", "p2ry12", "marco", "fybb", "csf1ra",
  "csf1rb", "ptprc", "rraga", "havcr1", "slc7a7",
  "ptpn6", "lyn", "spi1b", "c1qa", "c1qb", "c1qc",
  "csf2rb", "cd68", "LOC103034161", "spi1a",
  "LOC125803969", "myo1f",
  "epor", "hbbe1.1", "klf1", "lyl1", "myo1g",
  "grna", "ctsba", "ctsd", "ctsz", "lyz",
  "rac2", "ncf1", "ncf2", "ncf4", "ctsk",
  "rubcnl", "bin2b", "ctsc", "ctsh",
  "mafba", "runx1", "lcp1", "coro1a", "irf8",
  "irf5", "cd4-1", "cd8a", "cd8b", "il10ra",
  "tnfsf11", "il6", "il6r",
  
  # Ependymal - cilliated
  "dnah7", "dnah9", "tekt1", "tekt2", "tekt3",
  "cfap61", "cfap52", "cfap53", "foxj1b", "dnajc22",
  "tppp3", "ttc21a", "ccdc78", "dnah1", "dnah2",
  "dnah3", "dnah5l", "dnai2b", "dnai4",
  "cfap52", "cfap53", "cfap61", "cfap73",
  "cfap91", "cfap99", "cfap276", "spag6",
  "spag17", "spef1", "rsph3", "ccdc33",
  "ccdc40", "ccdc96", "ccdc146", "kif9"
)

other_markers <- c(
  "cldn5b", "vtnb", "acta2", "myh11a", 
  "col1a1a", "col2a1a", "epcam", "kdrl", "cldn5a", "pld1a", "mcamb", "kcne4", "rasl12",
  "slc47a2.1", "slc4a7", "slc5a6a", "slc16a9a", "col8a1a", "slc13a4", "soul5", "igfbp2a", 
  "ndufa4l2a","kcnj8", "rgs5a", "abcc9", "col9a2", "matn1",
  
  # Chondrocyte markers
  "trpv4", "adamts3", "ddr2a", "ddr2b", "papss2a", "papss2b", "vit", "ptger1b", "itga10",
  "itga11a", "tgfb5", "fgfr3", "fgfrl1b", "pdgfrl",

  # Core ECM and cartilage structure
  "acana", "acanb", "col2a1a", "col9a1a", "col9a2", "col9a3",
  "col11a1a", "col11a2", "col27a1a", "col27a1b", "matn1", "matn4", "epyc",
    
  # Cartilage regulators and modifiers
  "sox9a", "sox5", "sox6", "fgfr3", "fgfrl1b", "fgf2", "tgfb5",
    
  # Chondroitin synthesis and sulfation
  "papss2a", "chst3a", "chst3b", "chst7", "has3", "slc35d1a", "csgalnact1a", "xylt1",
    
  # Matrix crosslinking and assembly
  "loxl3a", "loxl3b", "serpinh1b", "snorc", "dpp4", "cthrc1b", "frzb",
 
  # Skeletal muscle 
  ## actin_regulators
  "synpo2la", "synpo2lb", "smpx", "fhl2a", "coro6", "unc45b",

  ## sarcomeric_ion_channels
  "ryr1a", "atp2a1", "cacng1a", "cacng6a", "cacng6b", "slc8a1a",

  ## neuromuscular_junction
  "chrna1", "chrnd", "chrng", "rapsn", "dok7b", "musk",

  ## Sarcomere and contractile machinery
  "acta1a", "acta1b", "actc1a", "actc1b", "actn3a", "actn3b",
  "myom1a", "myom2a", "myom2b", "myoz1a", "myoz1b", "myoz2a", "myl1", "mylpfa", 
  "tnnt3a", "neb", "desma", "klhl31", "klhl43", "klhl41b", "trim63a", "trim63b",

  ## muscle_regulators
  "myod1", "myf6", "mef2ca", "smyd1a", "smyd1b", "myf5","sh3bgr", "rbfox1l", 
  "ankrd1a", "xirp2a", "xirp2b", "coro6",

  ## Calcium handling and excitation-contraction coupling
  "atp2a1", "atp2a1l", "casq1b", "casq2", "ryr3", "cacna1ia", "cacna1sb",
  "jph2", "srl", "hrc", "plpp7a",

  ## Energy metabolism
  "ckma", "ckmb", "pfkmb", "aldoab", "gapdh", "pgam2", "pkmb", "gamt", "mdh1ab", "pdia8", "atp5f1b",

  # Endothelial markers
  "cdh5", "pecam1a", "pecam1b", "tie1", "tek", "flt1", "kdrl", "ptprb", "esama", "eng",
  "vwf","plvapb",

  ## Tight junction & BBB-related
  "cldn5a", "cldn5b", "jam2a", "podxl",

  ## Transporters / BBB-associated
  "slc2a1a", "slco1c1", "slc7a5", "abcg2a", "abcb4",
  
  ## Angiogenic transcription factors & signaling
  "sox7", "egfl7", "dll4",  "notchl", "lyve1a", "gpr182", "gpr183a", 

  ## Collagens
  "col6a1", "col6a2", "col6a4a", "col5a1", "col5a2a", "col10a1a",
  "col11a1b", "col12a1a", "col12a1b", "col21a1", "col22a1", "col8a2",

  # Epidermal 
  ## basal keratinocytes
  "tp63","krt5", "krt15", 
  "itgb4", "col17a1a", "col7a1", "col7a1l", 
  "lamc2", "lama3", "lamc1","cd44a", "fn1a",                      
  "barx2", "msx2a", "dlx3b",               

  ## Structural keratinocyte/epidermal markers
  "krt5", "krt15", "col7a1", "col7a1l", "col17a1a",

  ## Epithelial junction/adhesion
  "cldn1", "cldni", "cd44a", "itgb4", "itga6b", "itga3b", "lamc2", "lama5", "lamb3", "lamb4",

  ## Developmental regulators
  "tfap2c", "foxn1", "dlx3b", "shhb", "wnt3a", "wnt9b", "wnt10a",

  ## Epithelial signaling and remodeling
  "fgfbp1a", "adamts12", "mmp28", "mmp30", "postnb", "postna",

  ## Cell surface / periderm-specific
  "upk1a", "upk3b", "plp2a", "cxcl8a", "aqp3a", "aqp3b", "rbp4", "gpr84",
  "grhl1", "grhl2a","anxa1a", "anxa5a",         
  "tfap2c", "sp6", "arl13b", "odc1",              

  # Perivascular / Meningeal / Mesenchymal-like markers
  "aebp1a", "mfap2", "pmp22a", "tagln", "cxcl12a", "cd34", "angptl7", "plpp3",
  "abi3bpb", "phldb2a", "twist2", "twist3", "prrx1a", "prrx1b", "alx4a", "runx2a", 
    
  ## mural lymphatic endothelial cells (muLECs)  
  "epd","prox1a", "lyve1b", "mrc1a",  "slc13a4", "soul5",  "igfbp2a", "apof",
  "ggctb",
    
  ## pericytes  
  "cd248a", "pdgfrb", "notch3", "ndufa4l2a", "kcnj8", "rgs5a", "abcc9", "kcne4",
  "hey1", "hey2", "tcf12", "tcf15", "col4a2", "myo1b",  
    
  ## vascular smooth muscle cells (vSMC)  
  "acta2", "pdgfrb", "myh11a", "pln", "mcama", "mcamb",
    
  ## Meningeal Fibroblasts 
  "lum", "dcn", "col1a1a", "fn1a", "postnb", "klf2a", "klf15", "six1a", "slc47a2.1", 
  "slc4a7", "slc5a6a", "slc16a9a", "col8a1a", "aldh1a2", "col5a1", "meox2b", "tcf21",
  "pdgfra", "twist1a",
    
  ## Adhesion, basal lamina
  "cdh26.1", "cd44a", "cldn1", "epcam",
  "itga3b", "itga6b", "itgb4", "lamc2", "lama5", "lamb3", "lamb4",
  "col17a1a", "col7a1", "col4a4", "col4a5", "col4a6",
    
  ## Epidermal identity
  "krt5", "krt15", "grhl2a", "foxn1", "sp6", "tp63", "wnt10a", "fgf24", "barx2", "dlx3b",

  # Epithelial
  "grhl1", "cd109", "cdx1b", "ehf", "elf3", "ace2", "esrp1", "esrp2", "gjb9a", "vgll1",
  "ttr", "asl", "cldn1",
    
  ## tuft cells
  "pou2f3", "grhl1", "cd109", "cdx1b", "ehf", "elf3",
  "cldn1", "gjb9a", "esrp1", "esrp2",
  "vgll1", "ttr", "asl", "ace2",
   
  ## Choroid plexus epithelium
  "ttr",  "klf2a", "cfd", "slc4a5a", "slc4a5b", "cxcl12a", "foxj1a", "frzb",
  "otop1", "foxi1",  "foxa3", "hnf1ba", "hnf1a", "hnf4a", "otog", "apoa1a" , "meis2b",
  "crabp1a",
   
  ## Hair cells
  # Core hair cell bundle & stereocilia genes
  "pcdh15b", "ush1ga", "espn", "ptprq", "tmc2b", "myo15aa", "myo15ab",
  
  # Inner ear / lateral line specificity
  "myo6b", "myo7bb", "cabp2b", "pjvk", "sytl2b", 
  
  # Supporting transcription factors and regulators
  "gfi1ab", "gfi1aa", "foxj1b", "srrm4",
  
  # Ion regulation
   "atp1b2b", "scn12aa",
  
  # Cell adhesion / epithelial
  "cdhr2", "cdhr5b", "anxa5a", "espnla",

  # Liver
  "hnf4a", "hnf4b", "hnf4g","hnf1a", "hnf1ba", "hnf1bb",
  "pck1", "gys2", "fabp10a", "fabp2","tat", "aldob", "arg1", "pklr", "bhmt",
  "gpx3", "cdo1", "glud1a", "glud1b","apoa1b", "apoa4a", "apoba", "apoc1", "apoc2",
  "apom", "c4b", "c5", "c8g", "c9", "f2", "f5", "f7", "f10", "fgg", "fgb", "kng1",
  "serpinc1", "serpind1", "plg", "abcb11a", "abcb11b","slc2a2", "slc22a6l", "slc13a2",
  "slc22a13b", "slc7a7", "slc5a1","soat2", "dgat1a", "dgat2", "mogat2", "mogat3b",
  "nr1h4", "pparg", "pparda", "nr5a5", "fgf19", "ttr", "vtna", "klf15", "kl",
  "selenop2","onecut2"
)
```
### score cell type by top marker gene expression
default bin size = 24
```{r}
lSeq <- AddModuleScore(lSeq, features = list(neuronal_markers), name = "NeuronScore")
lSeq <- AddModuleScore(lSeq, features = list(glial_markers), name = "GliaScore")
lSeq <- AddModuleScore(lSeq, features = list(other_markers), name = "OtherScore")

# Classify by max score
lSeq$cell_type <- apply(
  lSeq@meta.data[, c("NeuronScore1", "GliaScore1", "OtherScore1")],
  1,
  function(x) c("Neuron", "Glia", "Other")[which.max(x)]
)
```
### visualize
```{r}
# Make a table of cell type per cluster
cell_type_cluster <- table(lSeq$seurat_clusters, lSeq$cell_type)

# Convert to data frame
df <- as.data.frame(cell_type_cluster)
colnames(df) <- c("Cluster", "CellType", "Count")

# Bar plot: side-by-side
ggplot(df, aes(x = Cluster, y = Count, fill = CellType)) +
  geom_bar(stat = "identity", position = "dodge") +
  ylab("Cell Count") +
  xlab("Seurat Cluster") +
  theme_minimal() +
  ggtitle("Cell Type Counts per Cluster") #+
 #theme(axis.text.x = element_text(angle = 45, hjust = 1))

#ggsave("lSeqMod.pdf", width = 12, height = 8, dpi = 300)
```

### Assign labels by cluster
```{r}
cluster_labels <- lSeq@meta.data %>%
  group_by(seurat_clusters, cell_type) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(seurat_clusters) %>%
  slice_max(n, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(seurat_clusters = as.numeric(seurat_clusters)) %>%  # now safely convert to numeric
  arrange(seurat_clusters)  # sort numerically

#print(cluster_labels)
```

```{r}
Idents(lSeq) <- lSeq$seurat_clusters
```

```{r}
new.cluster.ids <- cluster_labels$cell_type
names(new.cluster.ids) <- levels(lSeq)
lSeq <- RenameIdents(lSeq, new.cluster.ids)
lSeq$broadcelltypes <- Idents(lSeq)
```

```{r}
# sanity check
table(lSeq$seurat_clusters, lSeq$broadcelltypes)
```

### plot to check

```{r}
DimPlot(lSeq, reduction = "umap", group.by = "seurat_clusters", label = TRUE)
```
```{r}
DimPlot(lSeq, reduction = "umap", group.by = "broadcelltypes", label = TRUE)
```

## Refine annotations with DEGs
### Calculate degs
```{r}
pb <- AggregateExpression(
  lSeq,
  assays        = "RNA",
  group.by      = c("population", "sample_id", "seurat_clusters"),
  return.seurat = TRUE
)

cts <- GetAssayData(pb, assay = "RNA", layer = "counts")

coldata <- pb@meta.data %>%
  tibble::rownames_to_column("col_id") %>%     # rownames are the pseudo-bulk column IDs
  transmute(
    col_id     = col_id,
    sample     = sample_id,
    Population = factor(population, levels = c("surface", "cave")),
    CellType   = seurat_clusters
  )
```
all clusters
```{r}
res <- run_all_celltype_deg_exports(
  cts = cts,
  coldata = coldata,
  padj_cut = 0.05,
  lfc_min = 0.5,
  positive_only = TRUE,
  out_csv = "WBseurat_clusters_MarkerDEGs.csv"
)
```
single cluster
```{r}
res <- deseq2_pb_annotation(
  cts              = cts,     # pseudobulk count matrix (genes x columns)
  coldata          = coldata,     # pseudobulk metadata
  target_celltype  = "0"
)

format_and_write_annotation_csv(
  res_df = res$res_all,
  neuro_cluster_label = "WB0",
  padj_cut = 0.05,
  top_n = 500
)
```

### Refine annotations by cluster
```{r}
rm(non_cells)
# Get cluster cells
non_cells <- WhichCells(lSeq, idents = c("47"))

## Reassign identities
current_idents <- Idents(lSeq)
current_idents[non_cells] <- "Glia"
Idents(lSeq) <- current_idents
lSeq$finalbroadcelltypes <- Idents(lSeq) 
```

Check cluster on UMAP
```{r}
Idents(lSeq) <- lSeq$seurat_clusters

cluster_to_highlight <- "35"

DimPlot(lSeq,
  group.by = "seurat_clusters",
  cols = c("grey80", "red"),
  cells.highlight = WhichCells(lSeq, idents = 35),
  label = FALSE
) +
  ggtitle(paste("Highlighted cluster", cluster_to_highlight)) + NoLegend()

#chang idents back to avoid confusion
Idents(lSeq) <- lSeq$finalbroadcelltypes
```
## Refine annotations by subclustering original clusters
### subset data
```{r}
Idents(lSeq) <- lSeq$seurat_clusters
lSeqSub <- subset(x = lSeq, idents = c("20"))
Idents(lSeq) <- lSeq$finalbroadcelltypes
```

### reprocess subset
```{r}
lSeqSub <- FindNeighbors(lSeqSub, dims = 1:20, reduction = "pca")
lSeqSub <- FindClusters(lSeqSub, resolution =c(0.25, 0.5, 0.75))
```

### determine resolution
```{r}
#clustering tree
clustree(lSeqSub@meta.data, prefix = "RNA_snn_res.0", exprs = "data", node_colour = "sc3_stability",  
         node_text_colour = "white",show_axis=TRUE) + theme(legend.position = "bottom")
```

```{r}
#choose resolution according to clustree
Idents(lSeqSub) <- lSeqSub@meta.data[["RNA_snn_res.0.75"]]
lSeqSub <- RunUMAP(lSeqSub, dims = 1:20, reduction = "pca")
```

### Module scoring by subset cluster
```{r}
# default bin size is 24 - sometimes too large for smaller subclusters
binnum = 20

lSeqSub <- AddModuleScore(lSeqSub, features = list(neuronal_markers), name = "NeuronScore", nbin = binnum)
lSeqSub <- AddModuleScore(lSeqSub, features = list(glial_markers), name = "GliaScore", nbin = binnum)
lSeqSub <- AddModuleScore(lSeqSub, features = list(other_markers), name = "OtherScore", nbin = binnum)

# Classify by max score
lSeqSub$cell_type <- apply(
  lSeqSub@meta.data[, c("NeuronScore1", "GliaScore1", "OtherScore1")],
  1,
  function(x) c("Neuron", "Glia", "Other")[which.max(x)]
)
```

```{r}
# Make a table of cell type per cluster
cell_type_cluster <- table(lSeqSub$seurat_clusters, lSeqSub$cell_type)

# Convert to data frame
df <- as.data.frame(cell_type_cluster)
colnames(df) <- c("Cluster", "CellType", "Count")

# Bar plot: side-by-side
ggplot(df, aes(x = Cluster, y = Count, fill = CellType)) +
  geom_bar(stat = "identity", position = "dodge") +
  ylab("Cell Count") +
  xlab("Seurat Cluster") +
  theme_minimal() +
  ggtitle("Cell Type Counts per Cluster") 
 # theme(axis.text.x = element_text(angle = 45, hjust = 1))
```



### Marker genes by subset 
### UMAP distribution
```{r}
#DimPlot acording to cluster identity
DimPlot(object = (lSeqSub),label=TRUE,  pt.size = 0.5) +
  theme(axis.title.x=element_text(size=12),
        axis.title.y=element_text(size=12))
```
### Subcluster DEGs
#### sample level
```{r}
pb <- AggregateExpression(
  lSeqSub,
  assays        = "RNA",
  group.by      = c("population", "sample_id", "seurat_clusters"),
  return.seurat = TRUE
)

cts <- GetAssayData(pb, assay = "RNA", layer = "counts")

coldata <- pb@meta.data %>%
  tibble::rownames_to_column("col_id") %>%     # rownames are the pseudo-bulk column IDs
  transmute(
    col_id     = col_id,
    sample     = sample_id,
    Population = factor(population, levels = c("surface", "cave")),
    CellType   = seurat_clusters
  )
```
all clusters
```{r}
res <- run_all_celltype_deg_exports(
  cts = cts,
  coldata = coldata,
  padj_cut = 0.05,
  lfc_min = 0.5,
  positive_only = TRUE,
  out_csv = "Subcluster_MarkerDEGs.csv"
)
```


### re-assign idents by subcluster
```{r}
rm(non_cells)
# Get cell names in cluster x of subcluster
non_cells <- WhichCells(lSeqSub, idents = c("7"))

# make sure subset cell names match the full object
non_cells <- intersect(non_cells, colnames(lSeq))

# Reassign identities
current_idents <- Idents(lSeq)
current_idents[non_cells] <- "Glia"
Idents(lSeq) <- current_idents
```
## Check annotations against original cluster
```{r}

lSeq$finalbroadcelltypes <- Idents(lSeq)

# Get a table of cluster vs cell_type
tab <-table(lSeq$seurat_clusters,lSeq$finalbroadcelltypes)
tab
```
# . Whole brain analysis using major cell type group annotations
## UMAP figures
### All initial clusters
```{r}
p_final <- mini_axis_umap(
  lSeq,
  reduction = "umap",
  group.by = "seurat_clusters",
  label = TRUE,
  shuffle = TRUE,
  alpha = 0.8,
  pt.size = 0.5,
  axis_frac = 0.08,
  pad_frac  = 0.03
)

p_final
#ggsave("lSeqUmapNoCoords.pdf", p_final, width = 10, height = 8, dpi = 300)
```

### Major cell type groups
```{r}
#DimPlot acording to cluster identity
DimPlot(object = (lSeq), label=TRUE,  pt.size = 0.5,
         cols = c("goldenrod2","slateblue2", "lightcoral")) +
  theme(axis.title.x=element_text(size=12),
        axis.title.y=element_text(size=12))

#ggsave("WB.pdf", width = 10, height = 8, dpi = 300)
```

#### Dot plot
```{r}
DotPlot(lSeq, features = c(
  
 # Neuron
  "elavl3", "elavl4", "kcnc3a", "cacna1aa","rbfox3a", "snap25a",
  
  
 # Glia
 "msi1b", "qki2", "fabp7a", "glula", "slc1a2b", "kcnj10a", 
  
 # Other  
  "slc38a5b", "sparc", "ahnak", "mcamb","col1a1a","col1a2"
 
    )
) + coord_flip() + 
  scale_color_gradient(
    low = "floralwhite",
    high = "#4d0099",
    oob = scales::squish,
    limits = c(0,1),
    name = "avg.exp"
  ) +
  theme_bw(base_size = 12) 

#ggsave("lSeqMarksDot.pdf", dpi = 300)
```

#### Other
```{r}
FeaturePlot(lSeq, features = c("slc38a5b", "sparc", "ahnak"),  max.cutoff = 3, cols = c("grey",
    "lightcoral"), reduction = "umap", pt.size = 0.5)

#ggsave("featureOther.pdf", width = 10, height = 8, dpi = 300)
```

#### Neuron
```{r}
FeaturePlot(lSeq, features = c("elavl4", "kcnc3a", "cacna1aa"),  max.cutoff = 3, cols = c("grey",
    "goldenrod2"), reduction = "umap", pt.size = 0.5)

#ggsave("featureNeuron.pdf", width = 10, height = 8, dpi = 300)
```
#### Glia
```{r}
FeaturePlot(lSeq, features = c("msi1b", "qki2", "ptn"),  max.cutoff = 3, cols = c("grey",
    "slateblue2"), reduction = "umap", pt.size = 0.5)

#ggsave("featureGlia.pdf", width = 10, height = 8, dpi = 300)
```




