---
title: "Gene Check"
output: 
  html_notebook: 
    toc: true
    theme: cerulean
    number_sections: true
---
This notebook checks for the expression of a target gene or target gene set within defined Astyanax mexicanus cell types.

# . Environment
## Load libraries
```{r}
# Seurat
library(Seurat)

# DEG analysis
library(DESeq2)

#plotting
library(ggplot2)
library(ggrepel) # label volcano

#data management
library(dplyr)
```

## Define functions
surface vs cave 
```{r}
deseq2_pb_celltype <- function(cts, coldata, celltype,
                               fdr_explore = 0.20,
                               min_total_count = 10,
                               min_samples = 2,
                               shrink_type = c("apeglm", "ashr", "normal")) {

  suppressPackageStartupMessages({
    library(Matrix)
    library(DESeq2)
    library(dplyr)
    library(tibble)
  })

  shrink_type <- match.arg(shrink_type)

  stopifnot(is.matrix(cts) || inherits(cts, "dgCMatrix"))
  stopifnot(all(c("col_id","sample","Population","CellType") %in% colnames(coldata)))

  # Coerce to plain data.frame to avoid Rle/list columns breaking dplyr verbs
  coldata <- as.data.frame(coldata, stringsAsFactors = FALSE)

  coldata$col_id     <- as.character(coldata$col_id)
  coldata$sample     <- as.character(coldata$sample)
  coldata$Population <- as.character(coldata$Population)
  coldata$CellType   <- as.character(coldata$CellType)

  # Align count columns
  stopifnot(all(coldata$col_id %in% colnames(cts)))
  cts <- cts[, coldata$col_id, drop = FALSE]

  # Subset to the requested cell type (already pseudobulked columns)
  cd_ct <- coldata[coldata$CellType == celltype, , drop = FALSE]
  if (nrow(cd_ct) == 0) stop("No pseudobulk samples for celltype: ", celltype)

  cts_ct <- cts[, cd_ct$col_id, drop = FALSE]

  # Prefilter genes
  keep <- Matrix::rowSums(cts_ct >= min_total_count) >= min_samples
  cts_ct <- cts_ct[keep, , drop = FALSE]
  if (nrow(cts_ct) == 0) stop("All genes filtered out; relax thresholds.")

  # Build DESeq2 colData and reorder to match columns
  cd_ct2 <- unique(cd_ct[, c("col_id", "sample", "Population")])
  cd_ct2 <- cd_ct2[match(colnames(cts_ct), cd_ct2$col_id), , drop = FALSE]
  stopifnot(identical(colnames(cts_ct), cd_ct2$col_id))
  rownames(cd_ct2) <- cd_ct2$col_id

  # DESeq2
  dds <- DESeqDataSetFromMatrix(
    countData = round(cts_ct),
    colData   = cd_ct2,
    design    = ~ Population
  )

  dds$Population <- factor(dds$Population)
  if (!("surface" %in% levels(dds$Population))) {
    stop("Population must include 'surface'. Found: ", paste(levels(dds$Population), collapse = ", "))
  }
  if (length(levels(dds$Population)) != 2) {
    stop("This function assumes exactly 2 Population levels. Found: ", paste(levels(dds$Population), collapse = ", "))
  }

  dds$Population <- relevel(dds$Population, ref = "surface")
  dds <- DESeq(dds)

  alt <- setdiff(levels(dds$Population), "surface")[1]
  coef_name <- paste0("Population_", make.names(alt), "_vs_surface")
  if (!coef_name %in% resultsNames(dds)) {
    coef_name <- resultsNames(dds)[grepl("^Population_", resultsNames(dds))][1]
  }

  res <- results(dds, name = coef_name, alpha = fdr_explore) |>
    as.data.frame() |>
    tibble::rownames_to_column("gene")

  res_shr <- tryCatch({
    shr <- lfcShrink(dds, coef = coef_name, type = shrink_type)
    as.data.frame(shr) |> tibble::rownames_to_column("gene")
  }, error = function(e) {
    shr <- lfcShrink(dds, coef = coef_name, type = "normal")
    as.data.frame(shr) |> tibble::rownames_to_column("gene")
  })

  res_merged <- res |>
    dplyr::left_join(
      res_shr |> dplyr::select(gene, log2FoldChange) |> dplyr::rename(log2FoldChange_shrunk = log2FoldChange),
      by = "gene"
    ) |>
    dplyr::arrange(padj, dplyr::desc(abs(log2FoldChange_shrunk)))

  res_celltype <- res_merged |>
    dplyr::filter(!is.na(padj), padj < fdr_explore)

  list(
    dds          = dds,
    res_celltype = res_celltype,
    res_all      = res_merged,
    celltype     = celltype,
    coef_name    = coef_name,
    fdr_explore  = fdr_explore
  )
}
```

```{r}
run_target_pb_by_celltype <- function(
  cts,
  coldata,
  celltypes,
  target_genes,
  padj_cut = 0.05,
  fdr_explore = 0.20,
  min_total_count = 10,
  min_samples = 2,
  shrink_type = "apeglm"
) {
  suppressPackageStartupMessages({
    library(dplyr)
    library(purrr)
    library(tibble)
  })

  map_dfr(celltypes, function(ct) {

    message("Running: ", ct)

    out <- deseq2_pb_celltype(
      cts = cts,
      coldata = coldata,
      celltype = ct,
      fdr_explore = fdr_explore,
      min_total_count = min_total_count,
      min_samples = min_samples,
      shrink_type = shrink_type
    )

    res <- out$res_all

    if (!"log2FoldChange_shrunk" %in% colnames(res)) {
      res$log2FoldChange_shrunk <- res$log2FoldChange
    }

    res %>%
      filter(gene %in% target_genes) %>%
      mutate(
        celltype = ct,
        direction = case_when(
          is.na(padj) ~ "Not tested / NA",
          padj > padj_cut ~ "Not significant",
          log2FoldChange_shrunk > 0 ~ "Up in Cave",
          log2FoldChange_shrunk < 0 ~ "Up in Surface",
          TRUE ~ "No change"
        ),
        significant = !is.na(padj) & padj <= padj_cut,
        neglog10_padj = -log10(padj)
      ) %>%
      select(
        celltype, gene, direction, significant,
        padj, log2FoldChange, log2FoldChange_shrunk,
        baseMean, neglog10_padj,
        everything()
      )
  })
}
```

volcano
```{r}
make_target_volcano <- function(res, target_genes, title = "") {
  library(dplyr)
  library(ggplot2)
  library(ggrepel)

  if (!"log2FoldChange_shrunk" %in% colnames(res)) {
    res$log2FoldChange_shrunk <- res$log2FoldChange
  }

  de.volc <- res %>%
    mutate(
      logP = -log10(padj),
      sig = case_when(
        padj <= 0.05 & log2FoldChange_shrunk > 0 ~ "Up in Cave",
        padj <= 0.05 & log2FoldChange_shrunk < 0 ~ "Up in Surface",
        TRUE ~ "Not sig"
      ),
      label = ifelse(gene %in% target_genes, gene, NA)
    )

  ggplot(de.volc, aes(x = log2FoldChange_shrunk, y = logP, color = sig)) +
    geom_point(alpha = 0.6, size = 1.2) +
    geom_text_repel(
      aes(label = label),
      color = "black",
      size = 3,
      na.rm = TRUE,
      min.segment.length = 0
    ) +
    scale_color_manual(values = c(
      "Up in Cave" = "palevioletred",
      "Up in Surface" = "darkseagreen3",
      "Not sig" = "lightgray"
    )) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray50") +
    theme_minimal() +
    labs(
      title = title,
      x = "log2FC Cave vs Surface",
      y = "-log10 adjusted p-value"
    )
}
```

# . Data
## load
```{r}
#whole brain
lSeq <-readRDS("lSeq.rds")

#neurons
lSeqNeuroLB <-readRDS("lSeqNeuroLB.rds")

#glia
lSeqGliaGen <-readRDS("lSeqGliaGen.rds")

#other
lSeqNonLB <-readRDS("lSeqNonLB.rds")
```

# . Target Genes 
## Check for genes in data set
example genes of interest based on zebrafish names
```{r}
DotPlot(lSeq, features = c(
  
  #mch
  "pmch", "pmchl", "pmch1", "pmch2",
 
  
  #mchr
  "mchr1a", "mchr1b", "mchr2", "mchr2a",
  
  #Adenosine recepotors
  "adora1a", "adora1b",
  "adora2aa", "adora2ab", "adodra2a.1",
  "adora2b", "adora3", "adora3a", "adora3a.1"
  
)) + RotatedAxis()
```
actually annotated genes
```{r}
DotPlot(lSeq, features = c(
  
  #mch
  "pmch", "pmchl",
  
  #mchr
  "mchr1a", "mchr1b", "mchr2a", 
  
  #Adenosine recepotors
  "adora1b",
  "adora2ab", 
  "adora2b"
  
)) + RotatedAxis()
```

# . Target gene DEGs
```{r}
target_genes <- c(
  "pmch", "pmchl",
  "mchr1a", "mchr1b", "mchr2a",
  "adora1b", "adora2ab", "adora2b"
)
```

## Whole brain
```{r}
pb <- AggregateExpression(
  lSeq,
  assays        = "RNA",
  group.by      = c("population", "sample_id", "finalbroadcelltypes"),
  return.seurat = TRUE
)

cts <- GetAssayData(pb, assay = "RNA", layer = "counts")

coldata <- pb@meta.data %>%
  tibble::rownames_to_column("col_id") %>%     # rownames are the pseudo-bulk column IDs
  transmute(
    col_id     = col_id,
    sample     = sample_id,
    Population = factor(population, levels = c("surface", "cave")),
    CellType   = finalbroadcelltypes
  )
```

```{r}
group1_cts <- c("Neuron", "Glia", "Other")
DEGtargets_lSeq <- run_target_pb_by_celltype(cts, coldata, group1_cts, target_genes)
head(DEGtargets_lSeq)
#write.csv(DEGtargets_lSeq, "MCH_adenosine_target_DE_lSeq_broad.csv", row.names = FALSE)
```

## Neurons
```{r}
pb <- AggregateExpression(
  lSeqNeuroLB,
  assays        = "RNA",
  group.by      = c("population", "sample_id", "celltype"),
  return.seurat = TRUE
)

cts <- GetAssayData(pb, assay = "RNA", layer = "counts")

coldata <- pb@meta.data %>%
  tibble::rownames_to_column("col_id") %>%     # rownames are the pseudo-bulk column IDs
  transmute(
    col_id     = col_id,
    sample     = sample_id,
    Population = factor(population, levels = c("surface", "cave")),
    CellType   = celltype
  )
```

```{r}
group2_cts <- levels(Idents(lSeqNeuroLB))
DEGtargets_Neuro <- run_target_pb_by_celltype(cts, coldata, group2_cts, target_genes)
head(DEGtargets_Neuro)
write.csv(DEGtargets_Neuro, "MCH_adenosine_target_DE_Neuro_broad.csv", row.names = FALSE)
```
### i2 DEGs
```{r}
GlyHB_out <- deseq2_pb_celltype(cts, coldata, "i2", fdr_explore = 0.05)

res_GlyHB <- GlyHB_out$res_all
dds_GlyHB<- GlyHB_out$dds
```
```{r}
library(ggplot2)
library(ggrepel)
library(dplyr)

# res = your DESeq2 results table for ONE cell type

de.volc <- res_GlyHB  %>%
  mutate(
    logP = -log10(padj),

    sig = case_when(
      padj <= 0.05 & log2FoldChange_shrunk > 0 ~ "Up in Cave",
      padj <= 0.05 & log2FoldChange_shrunk < 0 ~ "Up in Surface",
      TRUE ~ "Not sig"
    ),

    label = ifelse(gene == "adora2b", gene, NA)
  )

p <- ggplot(de.volc,
            aes(x = log2FoldChange_shrunk,
                y = logP,
                color = sig)) +

  geom_point(
    position = position_jitter(width = 0, height = 0),
    alpha = 0.7,
    size = 1.5
  ) +

  # black outline circle around adora2b
  # outlined version of ONLY adora2b
  geom_point(
    data = subset(de.volc, gene == "adora2b"),
    aes(x = log2FoldChange_shrunk, y = logP),
    shape = 21,
    fill = NA,
    color = "black",
    stroke = 0.8,
    size = 1.5,   # EXACTLY same as main geom_point size
    inherit.aes = FALSE
  ) +

  geom_text_repel(
    aes(label = label),
    color = "black",
    size = 3,
    segment.color = "black",
    segment.size = 0.4,
    min.segment.length = 0,
    na.rm = TRUE
  ) +

  scale_color_manual(values = c(
    "Up in Cave" = "palevioletred",
    "Up in Surface" = "darkseagreen3",
    "Not sig" = "lightgray"
  )) +

  geom_vline(
    xintercept = c(-1, 1),
    linetype = "dashed",
    color = "gray50"
  ) +

  geom_hline(
    yintercept = -log10(0.05),
    linetype = "dashed",
    color = "gray50"
  ) +

  labs(
    title = "Volcano Plot: Cave vs Surface",
    x = "log2 Fold Change (Cave vs Surface)",
    y = "-log10 Adjusted p-value",
    color = "DE Category"
  ) +

  theme_minimal()

print(p)

ggsave("adora2b_volcano.pdf",
       p,
       width = 10,
       height = 8,
       dpi = 300)
```




## Glia
```{r}
pb <- AggregateExpression(
  lSeqGliaGen,
  assays        = "RNA",
  slot          = "counts",
  group.by      = c("population", "sample_id", "celltype"),
  return.seurat = TRUE
)

cts <- GetAssayData(pb, assay = "RNA", layer = "counts")

coldata <- pb@meta.data %>%
  tibble::rownames_to_column("col_id") %>%     # rownames are the pseudo-bulk column IDs
  transmute(
    col_id     = col_id,
    sample     = sample_id,
    Population = factor(population, levels = c("surface", "cave")),
    CellType   = celltype
  )
```

```{r}
group3_cts <- levels(Idents(lSeqGliaGen))
DEGtargets_Glia <- run_target_pb_by_celltype(cts, coldata, group3_cts, target_genes)
head(DEGtargets_Glia)
write.csv(DEGtargets_Glia, "MCH_adenosine_target_DE_Glia_broad.csv", row.names = FALSE)
```
## Other
```{r}
pb <- AggregateExpression(
  lSeqNonLB,
  assays        = "RNA",
  slot          = "counts",
  group.by      = c("population", "sample_id", "celltype"),
  return.seurat = TRUE
)

cts <- GetAssayData(pb, assay = "RNA", layer = "counts")

coldata <- pb@meta.data %>%
  tibble::rownames_to_column("col_id") %>%     # rownames are the pseudo-bulk column IDs
  transmute(
    col_id     = col_id,
    sample     = sample_id,
    Population = factor(population, levels = c("surface", "cave")),
    CellType   = celltype
  )
```

```{r}
group4_cts <- levels(Idents(lSeqNonLB))
DEGtargets_Other <- run_target_pb_by_celltype(cts, coldata, group4_cts, target_genes)
head(DEGtargets_Other)
write.csv(DEGtargets_Other, "MCH_adenosine_target_DE_Other_broad.csv", row.names = FALSE)
```

```{r}
group1_cts <- c("Neurons", "Glia", "Other")
group2_cts <- levels(Idents(lSeqNeurons))
group3_cts <- levels(Idents(lSeqGlia))
group4_cts <- levels(Idents(lSeqOther))

mch_group1 <- run_target_pb_by_celltype(cts, coldata, group1_cts, target_genes)
mch_group2 <- run_target_pb_by_celltype(cts, coldata, group2_cts, target_genes)
mch_group3 <- run_target_pb_by_celltype(cts, coldata, group3_cts, target_genes)
mch_group4 <- run_target_pb_by_celltype(cts, coldata, group4_cts, target_genes)

write.csv(mch_group1, "MCH_adenosine_target_DE_group1_broad.csv", row.names = FALSE)
write.csv(mch_group2, "MCH_adenosine_target_DE_group2_neurons.csv", row.names = FALSE)
write.csv(mch_group3, "MCH_adenosine_target_DE_group3_glia.csv", row.names = FALSE)
write.csv(mch_group4, "MCH_adenosine_target_DE_group4_other.csv", row.names = FALSE)
```










